---
title: "Evidence-ranking matrix for disease-group findings"
description: "Ranks the strongest diet, mycobiome, and symptom differences across Study_group_new to prioritize candidates for future validation."
date: 2026-05-29
categories: [pre-specified, synthesis, validation]
author: "IBD Capstone team"
---

## Research question

Which **disease-group** (`Study_group_new`) findings are **strongest candidates for future validation**?

We synthesize tests already run across the capstone pipeline into a single **evidence-ranking matrix**, ordered by a transparent score combining p-value and effect size.

## Data and inputs

| Domain | Tests included | Primary inputs |
|--------|----------------|----------------|
| Mycobiome (beta) | PERMANOVA × taxonomic level | `data/processed/{phylum,family,genus,species}.csv` |
| Mycobiome (alpha) | Kruskal–Wallis × diversity metric | `data/intermediate/alpha_long.rds` |
| Diet | Kruskal–Wallis × nutrient | `data/processed/dietary_cleaned.xlsx` + mycobiome metadata |
| Symptoms | Kruskal–Wallis × symptom score | `data/processed/cleaned_characteristics.csv` + mycobiome metadata |

- **Comparison:** Active IBD vs Quiescent vs Non-IBD (`Study_group_new`).
- **Unit:** Sample-level for mycobiome; participant-level for diet and symptoms (diet/symptoms joined to mycobiome participants).
- **Not included here:** Continuous symptom–composition associations (Spearman/PERMANOVA on HBI), stratified nutrient–diversity Spearman, or unsupervised PCA clusters — see linked posts below.

```{r}
#| label: setup
#| include: false
library(dplyr)
library(tidyr)
library(knitr)
library(here)
# Anchor here to stats/ so paths work both via `quarto render` and interactive IDE sessions.
here::i_am("posts/2026-05-29-evidence-ranking-disease-group/index.qmd")
source(here::here("R", "evidence_ranking_helpers.R"))

findings <- collect_study_group_findings()
ranked <- rank_evidence(findings)

n_tests <- nrow(ranked)
n_sig <- sum(ranked$p < 0.05, na.rm = TRUE)
n_high <- sum(ranked$validation_tier == "High", na.rm = TRUE)
```

**Summary:** `r n_tests` study-group comparisons ranked; `r n_sig` with p < 0.05; `r n_high` tagged **High** priority.

## Methods

| Item | Choice |
|------|--------|
| Integration | Recompute or aggregate from processed/intermediate files (same sources as individual posts) |
| Beta diversity | PERMANOVA (`adonis2`), Bray–Curtis, 999 permutations; effect = R² |
| Alpha, diet, symptoms | Kruskal–Wallis; effect = η²[H] (`rstatix::kruskal_effsize`) |
| Evidence score | `−log₁₀(p) × (0.5 + 0.5 × effect_scaled)` with R² or η²[H] capped at 1 |
| Validation tier | **High:** p < 0.05 and score ≥ 75th percentile among significant; **Medium:** p < 0.05; **Suggestive:** p < 0.10; **Low:** otherwise |
| Multiplicity | Exploratory ranking; no global FDR across all rows |

```{r}
#| label: run-ranking
top_validation <- ranked |>
  filter(validation_tier %in% c("High", "Medium")) |>
  slice_head(n = 10)
```

## Results

### Full evidence-ranking matrix

```{r}
#| label: ranking-table
#| echo: false
ranked |>
  select(
    rank, validation_tier, domain, feature, test,
    effect_label, statistic, p, n, evidence_score, source_post
  ) |>
  knitr::kable(
    digits = 4,
    col.names = c(
      "Rank", "Tier", "Domain", "Feature", "Test",
      "Effect", "Statistic", "p", "n", "Evidence score", "Source post"
    )
  )
```

### Top candidates for future validation

```{r}
#| label: top-table
#| echo: false
top_validation |>
  select(rank, validation_tier, domain, feature, test, effect_label, p, evidence_score) |>
  knitr::kable(
    digits = 4,
    caption = "Highest-priority findings (High or Medium tier, top 10 by evidence score)"
  )
```

```{r}
#| label: top-summary
#| echo: false
#| results: asis
if (nrow(top_validation) == 0) {
  cat("No findings reached **High** or **Medium** validation tier at p < 0.05.\n")
} else {
  cat("**Top validation candidates:**\n\n")
  for (i in seq_len(min(5L, nrow(top_validation)))) {
    r <- top_validation[i, ]
    cat(sprintf(
      "%d. [%s] **%s** — %s (%s, %s, p = %.4f, score = %.2f)\n",
      r$rank, r$validation_tier, r$domain, r$feature,
      r$effect_label, r$test, r$p, r$evidence_score
    ))
  }
}
```

### By domain (mean evidence score)

```{r}
#| label: domain-summary-table
#| echo: false
ranked |>
  group_by(domain) |>
  summarise(
    n_tests = n(),
    n_sig = sum(p < 0.05, na.rm = TRUE),
    mean_score = mean(evidence_score, na.rm = TRUE),
    top_feature = feature[which.max(evidence_score)],
  ) |>
  knitr::kable(
    digits = 3,
    col.names = c("Domain", "Tests", "p < 0.05", "Mean evidence score", "Top feature")
  )
```

## Figure

```{r}
#| label: fig-evidence-heatmap
#| fig-cap: "Evidence scores for the top 20 study-group comparisons. Higher scores indicate stronger statistical support for validation."
#| fig-width: 9
#| fig-height: 8
#| fig-align: center

plot_evidence_heatmap(ranked, top_n = min(20L, nrow(ranked)))
```

```{r}
#| label: save-figure
#| include: false
dir.create(here("figures"), showWarnings = FALSE, recursive = TRUE)
ggsave(
  here("figures", "evidence-ranking-matrix.png"),
  plot_evidence_heatmap(ranked, top_n = min(20L, nrow(ranked))),
  width = 9,
  height = 8,
  dpi = 300
)
```

## Interpretation

The matrix **does not replace** individual analysis posts; it prioritizes which disease-group contrasts to carry forward (e.g. pre-registration, larger cohort, or experimental follow-up).

**High** tier findings combine statistical significance with relatively large R² (composition) or η²[H] (rank-based outcomes). **Low** tier rows may still be biologically interesting but need replication. Domains with few participants (especially diet and symptoms) should be interpreted cautiously.

### Related analyses (not in this matrix)

| Topic | Post |
|-------|------|
| Symptom × diversity (Spearman, HBI PERMANOVA) | `2026-05-22-symptoms-fungal-composition` |
| Nutrient × diversity (stratified Spearman) | `2026-05-29-nutrients-diversity-disease-group` |
| Diet + mycobiome PCA / clustering | `2026-05-29-diet-mycobiome-pca-clustering` |

## Reproducibility

- **Helpers:** `stats/R/evidence_ranking_helpers.R`
- **Render:** `quarto render stats` from repository root
