---
title: "Nutrients vs fungal diversity by disease group (Spearman)"
description: "Spearman correlations between dietary nutrients and alpha diversity, overall and stratified by study group (Active IBD, Quiescent, Non-IBD)."
date: 2026-05-29
categories: [pre-specified, alpha-diversity, diet, nutrients]
author: "IBD Capstone team"
---

## Research question

Are certain **nutrients** associated with **fungal alpha diversity**, and do those associations **differ across disease groups**?

**Null (overall):** No monotonic association between each nutrient and each alpha diversity metric across all samples.

**Null (stratified):** Within each study group, no monotonic association between nutrient intake and alpha diversity.

Comparing **overall vs group-specific** Spearman \(\rho\) is exploratory (small n per group); formal tests of correlation differences were not run.

## Data

| Role | Source |
|------|--------|
| Mycobiome (alpha) | `data/intermediate/alpha_long.rds` (`make wrangle`) |
| Diet | `data/processed/dietary_cleaned.xlsx` (`make dietary` on `data/raw/SYN_dietary data(ALL).csv`) |
| Disease group | `Study_group_new` on mycobiome metadata (from `SYN_MBI sample IDs meta.xlsx`) |

- **Unit of analysis:** Stool sample (16 samples, 12 participants); nutrients are **participant means** across diet diary days (same as `merge_files.R`).
- **Join key:** `Participant_ID` = `Participant ID (ESHA ID)` (normalized SYN codes).
- **Nutrients:** Pre-specified panel (fiber, tryptophan, calories, sweeteners, sugars, fats, omegas, vitamin D, folate); see helpers.
- **Inclusion:** Samples retained after `02_data_wrangling.R` removals.

```{r}
#| label: setup
#| include: false
library(dplyr)
library(tidyr)
library(knitr)
library(here)
# Anchor here to stats/ so paths work both via `quarto render` and interactive IDE sessions.
here::i_am("posts/2026-05-29-nutrients-diversity-disease-group/index.qmd")
source(here::here("R", "nutrient_association_helpers.R"))

sample_nutrients <- build_sample_nutrient_table()

n_samples <- nrow(sample_nutrients)
n_participants <- n_distinct(sample_nutrients$Participant_ID)
n_with_fiber <- sum(!is.na(sample_nutrients[["TotFib (g)"]]))
groups <- sort(unique(sample_nutrients$Study_group_new))
```

**Coverage:** `r n_samples` samples, `r n_participants` participants, `r n_with_fiber` samples with fiber data; study groups: `r paste(groups, collapse = ", ")`.

## Methods

| Item | Choice |
|------|--------|
| Test | Spearman rank correlation (`cor.test`, `method = "spearman"`) |
| Response | Shannon, Simpson, Chao1 |
| Predictors | Participant-mean nutrient intake (g, kcal, mg as recorded) |
| Stratification | `Study_group_new` (Active IBD, Quiescent, Non-IBD) + **Overall** |
| Multiple testing | Benjamini–Hochberg within each stratum (across nutrient × metric pairs) |
| Minimum n | Correlation only if ≥ 4 complete pairs |
| Significance | α = 0.05 |

```{r}
#| label: run-tests
spearman_results <- spearman_nutrient_alpha(sample_nutrients)
```

## Assumptions and diagnostics

- **Independence:** Up to two stool replicates per participant; dietary intake is duplicated across replicates. Exploratory only.
- **Spearman:** Monotonic association; no linearity assumption.
- **Stratified n:** Few samples per study group limits power; many strata will return `NA` correlations.

## Results

### Overall correlations

```{r}
#| label: spearman-overall-table
#| echo: false
spearman_results |>
  filter(stratum == "Overall") |>
  select(nutrient_label, metric, n, rho, p, p_adj) |>
  knitr::kable(
    digits = 4,
    col.names = c("Nutrient", "Alpha metric", "n", "Spearman rho", "p", "BH-adjusted p")
  )
```

### Stratified by study group

```{r}
#| label: spearman-stratified-table
#| echo: false
spearman_results |>
  filter(stratum != "Overall") |>
  select(stratum, nutrient_label, metric, n, rho, p, p_adj) |>
  knitr::kable(
    digits = 4,
    col.names = c("Study group", "Nutrient", "Alpha metric", "n", "Spearman rho", "p", "BH-adjusted p")
  )
```

```{r}
#| label: spearman-summary
#| echo: false
#| results: asis
sig <- spearman_results |>
  filter(!is.na(p_adj), p_adj < 0.05)

if (nrow(sig) == 0) {
  cat("**No nutrient × diversity pairs were significant after BH adjustment (q < 0.05) in any stratum.**\n")
} else {
  cat("**Significant after BH adjustment (q < 0.05):**\n\n")
  for (i in seq_len(nrow(sig))) {
    cat(sprintf(
      "- [%s] %s vs %s: rho = %.3f, q = %.4f (n = %d)\n",
      sig$stratum[i], sig$nutrient_label[i], sig$metric[i],
      sig$rho[i], sig$p_adj[i], sig$n[i]
    ))
  }
}

# Highlight fiber × Shannon across strata for interpretation
fiber_shannon <- spearman_results |>
  filter(nutrient == "TotFib (g)", metric == "Shannon") |>
  select(stratum, n, rho, p)

cat("\n\n**Total fiber × Shannon (all strata):**\n\n")
cat(paste(
  sprintf("%s: rho = %s (n = %d, p = %s)",
          fiber_shannon$stratum,
          ifelse(is.na(fiber_shannon$rho), "NA", sprintf("%.3f", fiber_shannon$rho)),
          fiber_shannon$n,
          ifelse(is.na(fiber_shannon$p), "NA", sprintf("%.4f", fiber_shannon$p))),
  collapse = "\n"
))
```

## Figures

### Total fiber vs Shannon diversity (by study group)

```{r}
#| label: fig-fiber-shannon
#| fig-cap: "Total fiber intake vs Shannon diversity. Points are stool samples; diet is participant-mean intake. Facets show study group."
#| fig-width: 8
#| fig-height: 5
#| fig-align: center

plot_nutrient_alpha_scatter(
  sample_nutrients,
  nutrient_col = "TotFib (g)",
  metric = "Shannon",
  facet_by = "Study_group_new"
)
```

### Tryptophan vs Shannon diversity (by study group)

```{r}
#| label: fig-trp-shannon
#| fig-cap: "Tryptophan intake vs Shannon diversity, faceted by study group."
#| fig-width: 8
#| fig-height: 5
#| fig-align: center

plot_nutrient_alpha_scatter(
  sample_nutrients,
  nutrient_col = "Trp (g)",
  metric = "Shannon",
  facet_by = "Study_group_new"
)
```

```{r}
#| label: save-figures
#| include: false
dir.create(here("figures"), showWarnings = FALSE, recursive = TRUE)
ggsave(
  here("figures", "nutrients-fiber-shannon-by-group.png"),
  plot_nutrient_alpha_scatter(
    sample_nutrients, "TotFib (g)", "Shannon", "Study_group_new"
  ),
  width = 8,
  height = 5,
  dpi = 300
)
ggsave(
  here("figures", "nutrients-trp-shannon-by-group.png"),
  plot_nutrient_alpha_scatter(
    sample_nutrients, "Trp (g)", "Shannon", "Study_group_new"
  ),
  width = 8,
  height = 5,
  dpi = 300
)
```

## Interpretation

Overall Spearman correlations pool all study groups; **stratified** correlations describe associations within Active IBD, Quiescent, and Non-IBD separately. A nutrient may correlate with diversity overall but not within groups (or vice versa) when group-specific slopes differ or sample sizes are small.

These results do **not** test whether two correlation coefficients differ significantly between groups; that would require larger n or dedicated methods (e.g. comparing dependent correlations).

## Reproducibility

- **Helpers:** `stats/R/nutrient_association_helpers.R`
- **Diet pipeline:** `src/diet/dietary.py`
- **Merge reference:** `src/merge_files.R` (`aggregate_dietary`)
- **Render:** `quarto render stats` from repository root
