---
title: "Dietary EDA"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

This notebook is a streamlined exploratory analysis of the dietary dataset for this capstone project. Its purpose is to summarize nutrient and food-group intake, show the main distributions in the cleaned dietary records, and provide a small set of interpretable comparisons that can help guide later analyses

Also serves as a readable first pass over the dietary variables before modeling or integration with the microbiome + characteristics data. The notebook uses `data/processed/dietary_cleaned.xlsx` as the main source for diet variables, and only uses `data/processed/merged.csv` when metadata such as `Study_group_new` or `Fiber_restriction` is actually needed (bottom cells)

# Libraries

```{r}
library(tidyverse)
library(readxl)
library(janitor)
library(scales)
library(patchwork)
library(knitr)
```

# Load Processed Data

```{r}
dietary_raw <- read_excel("../data/processed/dietary_cleaned.xlsx", sheet = "Data") |>
  clean_names()

meta_raw <- read_csv("../data/processed/merged.csv", show_col_types = FALSE) |>
  clean_names()

dietary_raw |>
  select(participant_id_esha_id, timepoint, day) |>
  head()
```

# Prep A Participant-Level Dietary Table

```{r}
id_cols <- c("participant_id_esha_id", "timepoint", "day")

diet_cols <- c(
  "wgt_g", "cals_kcal", "prot_g", "carb_g", "fat_g", "tot_fib_g", "water_g",
  "trp_g", "mono_fat_g", "alc_g",
  "mp_grain_oz_eq", "mp_veg_c_eq", "mp_fruit_c_eq", "mp_dairy_c_eq", "mp_prot_oz_eq"
)

metadata_df <- meta_raw |>
  select(participant_id, study_group_new, fiber_restriction) |>
  distinct() |>
  filter(!is.na(participant_id))

diet_df <- dietary_raw |>
  select(any_of(c(id_cols, diet_cols))) |>
  filter(
    !str_detect(coalesce(day, ""), regex("^average\\s*$", ignore_case = TRUE)),
    !str_detect(coalesce(day, ""), regex("%\\s*recommendation", ignore_case = TRUE))
  ) |>
  mutate(
    participant_id = str_extract(participant_id_esha_id, "SYN_\\d+"),
    across(
      any_of(diet_cols),
      ~ readr::parse_number(as.character(.x))
    )
  ) |>
  fill(participant_id, .direction = "down") |>
  filter(!is.na(participant_id)) |>
  group_by(participant_id) |>
  summarise(across(any_of(diet_cols), ~ mean(.x, na.rm = TRUE)), .groups = "drop") |>
  left_join(metadata_df, by = "participant_id") |>
  filter(!is.na(study_group_new)) |>
  mutate(
    study_group_new = factor(
      study_group_new,
      levels = c("Non-IBD", "Quiescent", "Active IBD")
    )
  )

glimpse(diet_df)
```

# Theme (makes the plots look nicer)

```{r}
theme_diet <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold"),
      axis.title = element_text(face = "bold"),
      axis.text.x = element_text(angle = 30, hjust = 1),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}
```

# Quick checks

```{r}
summary_tbl <- tibble(
  metric = c(
    "Participants",
    "Study groups",
    "Fiber restriction categories",
    "Diet variables used"
  ),
  value = c(
    n_distinct(diet_df$participant_id),
    n_distinct(diet_df$study_group_new),
    n_distinct(diet_df$fiber_restriction),
    length(diet_cols)
  )
)

kable(summary_tbl)
```

```{r}
missing_tbl <- diet_df |>
  summarise(across(where(is.numeric), ~ mean(is.na(.x)) * 100)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") |>
  arrange(desc(pct_missing))

kable(missing_tbl, digits = 1)
```

# Descriptive summary

```{r}
key_nutrients <- intersect(
  c("cals_kcal", "prot_g", "carb_g", "fat_g", "tot_fib_g", "trp_g"),
  names(diet_df)
)

diet_df |>
  select(study_group_new, any_of(key_nutrients)) |>
  pivot_longer(-study_group_new, names_to = "nutrient", values_to = "value") |>
  group_by(study_group_new, nutrient) |>
  summarise(
    n = sum(!is.na(value)),
    mean = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(across(c(mean, sd, median), round, 2)) |>
  kable()
```

# Distribution plots

```{r}
plot_vars <- intersect(
  c("cals_kcal", "prot_g", "carb_g", "fat_g", "tot_fib_g", "trp_g"),
  names(diet_df)
)

dist_plot <- diet_df |>
  select(any_of(plot_vars)) |>
  pivot_longer(everything(), names_to = "nutrient", values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 18, fill = "#4E79A7", color = "white", alpha = 0.9) +
  facet_wrap(~ nutrient, scales = "free") +
  labs(
    title = "Distributions of Key Dietary Variables",
    x = NULL,
    y = "Count"
  ) +
  theme_diet()

dist_plot
```

# Group comparisons

```{r}
box_vars <- intersect(
  c("tot_fib_g", "trp_g", "cals_kcal", "prot_g"),
  names(diet_df)
)

group_plot <- diet_df |>
  select(study_group_new, any_of(box_vars)) |>
  pivot_longer(-study_group_new, names_to = "nutrient", values_to = "value") |>
  ggplot(aes(x = study_group_new, y = value, fill = study_group_new)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.7) +
  geom_jitter(width = 0.15, alpha = 0.45, size = 1.8, show.legend = FALSE) +
  facet_wrap(~ nutrient, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Dietary Variables by Study Group",
    x = NULL,
    y = NULL,
    fill = "Study group"
  ) +
  theme_diet()

group_plot
```

```{r}
test_vars <- intersect(c("tot_fib_g", "trp_g"), names(diet_df))

kruskal_results <- map_dfr(test_vars, function(var) {
  test <- kruskal.test(reformulate("study_group_new", response = var), data = diet_df)

  tibble(
    nutrient = var,
    statistic = unname(test$statistic),
    p_value = test$p.value
  )
}) |>
  mutate(
    statistic = round(statistic, 3),
    p_value = signif(p_value, 3)
  )

kable(kruskal_results)
```

```{r}
pairwise_results <- map_dfr(test_vars, function(var) {
  pw <- pairwise.wilcox.test(
    x = diet_df[[var]],
    g = diet_df$study_group_new,
    p.adjust.method = "BH"
  )

  expand.grid(
    group_1 = rownames(pw$p.value),
    group_2 = colnames(pw$p.value),
    stringsAsFactors = FALSE
  ) |>
    as_tibble() |>
    mutate(
      p_value = map2_dbl(group_1, group_2, ~ pw$p.value[.x, .y])
    ) |>
    filter(!is.na(p_value)) |>
    mutate(nutrient = var, p_value = signif(p_value, 3))
})

kable(pairwise_results)
```

# Healthy Diet scores

## Food group score

```{r}
score_cols <- intersect(c(
  "mp_grain_oz_eq", "mp_veg_c_eq", "mp_fruit_c_eq", "mp_dairy_c_eq", "mp_prot_oz_eq"
), names(diet_df))

diet_df <- diet_df |>
  mutate(
    healthy_diet_score = rowSums(across(any_of(score_cols)), na.rm = TRUE)
  )

diet_df |>
  summarise(
    mean_score = mean(healthy_diet_score, na.rm = TRUE),
    sd_score = sd(healthy_diet_score, na.rm = TRUE),
    median_score = median(healthy_diet_score, na.rm = TRUE)
  ) |>
  mutate(across(everything(), round, 2)) |>
  kable()
```

## Mediterranean-Style Score (missing many food components)

```{r}
score_medians <- diet_df |>
  summarise(across(any_of(score_cols), median, na.rm = TRUE))

diet_df <- diet_df |>
  mutate(
    mono_fat_ratio = mono_fat_g / if_else(fat_g == 0, NA_real_, fat_g),
    mds = 0L
  )

if ("mp_veg_c_eq" %in% names(diet_df)) {
  diet_df <- diet_df |>
    mutate(mds = mds + if_else(mp_veg_c_eq >= score_medians$mp_veg_c_eq, 1L, 0L, missing = 0L))
}

if ("mp_fruit_c_eq" %in% names(diet_df)) {
  diet_df <- diet_df |>
    mutate(mds = mds + if_else(mp_fruit_c_eq >= score_medians$mp_fruit_c_eq, 1L, 0L, missing = 0L))
}

if ("mp_grain_oz_eq" %in% names(diet_df)) {
  diet_df <- diet_df |>
    mutate(mds = mds + if_else(mp_grain_oz_eq >= score_medians$mp_grain_oz_eq, 1L, 0L, missing = 0L))
}

if ("mp_dairy_c_eq" %in% names(diet_df)) {
  diet_df <- diet_df |>
    mutate(mds = mds + if_else(mp_dairy_c_eq < score_medians$mp_dairy_c_eq, 1L, 0L, missing = 0L))
}

diet_df <- diet_df |>
  mutate(
    mds = mds + if_else(mono_fat_ratio > 0.35, 1L, 0L, missing = 0L),
    mds = mds + if_else(alc_g > 0 & alc_g <= 14, 1L, 0L, missing = 0L)
  )

diet_df |>
  ggplot(aes(x = study_group_new, y = mds, fill = study_group_new)) +
  geom_boxplot(alpha = 0.85) +
  geom_jitter(width = 0.15, alpha = 0.45, size = 1.8, show.legend = FALSE) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Mediterranean-Style Diet Score by Study Group",
    x = NULL,
    y = "MDS"
  ) +
  theme_diet()
```

# Intake vs. Reference Targets

```{r}
baseline_targets <- c(
  mp_grain_oz_eq = 6.0,
  mp_veg_c_eq = 2.5,
  mp_fruit_c_eq = 2.0,
  mp_dairy_c_eq = 3.0,
  mp_prot_oz_eq = 5.5
)

baseline_targets <- baseline_targets[names(baseline_targets) %in% names(diet_df)]

comparison_df <- tibble(
  variable = names(baseline_targets),
  target = unname(baseline_targets),
  average_intake = map_dbl(names(baseline_targets), ~ mean(diet_df[[.x]], na.rm = TRUE))
) |>
  mutate(
    average_intake = round(average_intake, 2),
    gap = round(average_intake - target, 2)
  )

kable(comparison_df)
```

```{r}
comparison_df |>
  pivot_longer(c(average_intake, target), names_to = "type", values_to = "value") |>
  ggplot(aes(x = variable, y = value, fill = type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("average_intake" = "#4E79A7", "target" = "#F28E2B")) +
  labs(
    title = "Average Intake vs. Reference Targets",
    x = NULL,
    y = "Amount",
    fill = NULL
  ) +
  theme_diet()
```

```{r}
deficiency_df <- diet_df |>
  transmute(
    participant_id,
    across(
      all_of(names(baseline_targets)),
      ~ .x - baseline_targets[cur_column()],
      .names = "{.col}_deviation"
    )
  ) |>
  mutate(
    total_deficiency_score = rowSums(
      across(
        ends_with("_deviation"),
        ~ if_else(.x < 0, .x, 0)
      ),
      na.rm = TRUE
    )
  ) |>
  arrange(total_deficiency_score)

kable(head(deficiency_df, 10), digits = 2)
```

# Correlation plot

```{r}
corr_vars <- intersect(c(
  "cals_kcal", "prot_g", "carb_g", "fat_g", "tot_fib_g", "trp_g",
  "mp_grain_oz_eq", "mp_veg_c_eq", "mp_fruit_c_eq", "mp_dairy_c_eq", "mp_prot_oz_eq"
), names(diet_df))

cor_mat <- cor(diet_df[, corr_vars], use = "pairwise.complete.obs")

corr_long <- expand.grid(
  var_1 = rownames(cor_mat),
  var_2 = colnames(cor_mat),
  stringsAsFactors = FALSE
) |>
  as_tibble() |>
  mutate(
    corr = map2_dbl(var_1, var_2, ~ cor_mat[.x, .y])
  )

corr_long |>
  ggplot(aes(var_1, var_2, fill = corr)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#B07AA1",
    mid = "white",
    high = "#4E79A7",
    limits = c(-1, 1)
  ) +
  labs(
    title = "Correlation Matrix of Selected Dietary Variables",
    x = NULL,
    y = NULL,
    fill = "r"
  ) +
  theme_diet() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Notes 

- This notebook intentionally starts from `data/processed/dietary_cleaned.xlsx` instead of the raw dietary file
- The merged dataset is only used to attach metadata needed for grouped comparisons
- The participant-level table averages day-level dietary records within each participant
- Duplicated exploratory plots, quartile binning, and unfinished clustering steps from the Python notebook were removed to keep the analysis focused and readable
