---
title: "Symptom scores vs fungal composition (Spearman and PERMANOVA)"
description: "Survey symptom scores (HBI and related items) tested against alpha diversity (Spearman) and genus beta diversity (PERMANOVA on Bray–Curtis)."
date: 2026-05-22
categories: [pre-specified, alpha-diversity, beta-diversity, genus, symptoms]
author: "IBD Capstone team"
---

## Research question

Are **symptom scores** from participant surveys associated with **fungal composition** (alpha and beta diversity)?

**Null (Spearman):** No monotonic association between each symptom score and each alpha diversity metric.

**Null (PERMANOVA):** Genus-level community composition is not associated with the symptom score after accounting for Bray–Curtis distance structure.

## Data

| Role | Source |
|------|--------|
| Mycobiome (alpha) | `data/intermediate/alpha_long.rds` (`make wrangle`) |
| Mycobiome (genus beta) | `data/processed/genus.csv` (`make save`) |
| Survey / symptoms | `data/processed/cleaned_characteristics.csv` (`make characteristics` on `data/raw/SYN_Participant Characteristics(Sheet1).csv`) |

- **Unit of analysis:** Stool sample (16 samples, 12 participants with mycobiome data).
- **Join key:** `Participant_ID` (mycobiome) = `participant_id` (characteristics); one characteristics row per participant used when duplicates exist.
- **Symptom scores:** Numeric fields and ordinally coded items parsed to numbers (e.g. `Mild = 1`, `4\tSOME OF THE TIME` → 4). Primary focus: **Harvey–Bradshaw Index (HBI)**; additional exploratory symptoms in the correlation table.
- **Inclusion:** Samples retained after `02_data_wrangling.R` removals; PERMANOVA genus analysis uses samples with non-missing HBI.

```{r}
#| label: setup
#| include: false
library(dplyr)
library(tidyr)
library(knitr)
library(here)
# Anchor here to stats/ so paths work both via `quarto render` and interactive IDE sessions.
here::i_am("posts/2026-05-22-symptoms-fungal-composition/index.qmd")
source(here::here("R", "symptom_association_helpers.R"))

sample_symptoms <- build_sample_symptom_table()
genus_wide <- load_taxa_wide("genus") |>
  left_join(
    load_symptom_scores(),
    by = "Participant_ID"
  )

n_samples <- nrow(sample_symptoms)
n_participants <- n_distinct(sample_symptoms$Participant_ID)
n_hbi <- sum(!is.na(sample_symptoms$harvey_bradshaw_index))
```

**Coverage:** `r n_samples` samples, `r n_participants` participants; `r n_hbi` samples with non-missing HBI.

## Methods

| Item | Spearman (symptoms × alpha) | PERMANOVA (symptoms × taxa) |
|------|----------------------------|-----------------------------|
| Test | Spearman rank correlation (`cor.test`, `method = "spearman"`) | `vegan::adonis2` |
| Response | Shannon, Simpson, Chao1 | Genus Bray–Curtis dissimilarity |
| Predictor | Parsed symptom scores | Harvey–Bradshaw Index (continuous) |
| Multiple testing | Benjamini–Hochberg across all symptom × metric pairs | Exploratory single predictor |
| Permutations | — | 999 |
| Significance | α = 0.05 | α = 0.05 |

```{r}
#| label: run-tests
spearman_results <- spearman_symptom_alpha(sample_symptoms)

permanova_hbi <- run_permanova_symptom(
  genus_wide,
  prefix = TAXA_LEVELS$genus$prefix,
  symptom_col = "harvey_bradshaw_index"
)
```

## Assumptions and diagnostics

- **Independence:** Samples are not fully independent (up to two stool replicates per participant). Results are exploratory; consider participant-level sensitivity analyses if reporting formally.
- **Spearman:** Tests monotonic association; does not assume linearity or normality.
- **PERMANOVA:** Exchangeability under permutations; composition constrained to genus relative abundance (Bray–Curtis).
- **Missing symptoms:** HBI and other scores are not available for all participants (see coverage above).

## Results

### Spearman: symptom scores × alpha diversity

```{r}
#| label: spearman-table
#| echo: false
spearman_results |>
  select(symptom_label, metric, n, rho, p, p_adj) |>
  knitr::kable(
    digits = 4,
    col.names = c("Symptom", "Alpha metric", "n", "Spearman rho", "p", "BH-adjusted p")
  )
```

```{r}
#| label: spearman-summary
#| echo: false
#| results: asis
sig <- spearman_results |>
  filter(!is.na(p_adj), p_adj < 0.05)

if (nrow(sig) == 0) {
  cat("**No symptom × alpha pairs were significant after BH adjustment (q < 0.05).**\n")
} else {
  cat("**Significant after BH adjustment (q < 0.05):**\n\n")
  for (i in seq_len(nrow(sig))) {
    cat(sprintf(
      "- %s vs %s: rho = %.3f, q = %.4f (n = %d)\n",
      sig$symptom_label[i], sig$metric[i], sig$rho[i], sig$p_adj[i], sig$n[i]
    ))
  }
}
```

### PERMANOVA: genus composition ~ HBI

```{r}
#| label: permanova-table
#| echo: false
knitr::kable(
  permanova_to_df(permanova_hbi$permanova),
  digits = 4,
  col.names = c("Term", "Df", "Sum of squares", "R²", "F", "Pr(>F)")
)
```

```{r}
#| label: permanova-summary
#| echo: false
#| results: asis
r2 <- permanova_hbi$permanova$R2[1]
p <- permanova_hbi$permanova$`Pr(>F)`[1]
cat(sprintf(
  "**Harvey–Bradshaw Index** explained **%.1f%%** of variance in genus composition (R² = %.3f; pseudo-F = %.2f; p = %.4f, 999 permutations; n = %d samples with HBI).",
  100 * r2,
  r2,
  permanova_hbi$permanova$F[1],
  p,
  nrow(permanova_hbi$metadata)
))
```

## Figures

### HBI vs Shannon diversity (Spearman)

```{r}
#| label: fig-hbi-shannon
#| fig-cap: "Harvey–Bradshaw Index vs Shannon diversity by study group. Subtitle shows Spearman correlation across samples with complete data."
#| fig-width: 7
#| fig-height: 5
#| fig-align: center

plot_symptom_alpha_scatter(
  sample_symptoms,
  symptom_col = "harvey_bradshaw_index",
  metric = "Shannon"
)
```

### Genus PCoA coloured by HBI (PERMANOVA)

```{r}
#| label: fig-pcoa-hbi
#| fig-cap: "PCoA of genus Bray–Curtis dissimilarity; points coloured by Harvey–Bradshaw Index."
#| fig-width: 7
#| fig-height: 5.5
#| fig-align: center

plot_pcoa_symptom(
  permanova_hbi,
  TAXA_LEVELS$genus$label
)
```

```{r}
#| label: save-figures
#| include: false
dir.create(here("figures"), showWarnings = FALSE, recursive = TRUE)
ggsave(
  here("figures", "symptoms-hbi-shannon.png"),
  plot_symptom_alpha_scatter(sample_symptoms, "harvey_bradshaw_index", "Shannon"),
  width = 7,
  height = 5,
  dpi = 300
)
ggsave(
  here("figures", "symptoms-permanova-genus-hbi-pcoa.png"),
  plot_pcoa_symptom(permanova_hbi, TAXA_LEVELS$genus$label),
  width = 7,
  height = 5.5,
  dpi = 300
)
```

## Interpretation

Spearman correlations describe **pairwise monotonic links** between symptom scores and alpha metrics at the sample level; they do not adjust for repeated measures per participant. PERMANOVA tests whether **multivariate genus composition** covaries with HBI across samples; a non-significant p-value does not rule out associations with individual taxa or other symptom instruments.

Correlation among symptoms and small sample size limit power; treat findings as **exploratory** unless pre-registered.

## Reproducibility

- **Helpers:** `stats/R/symptom_association_helpers.R`, `stats/R/permanova_helpers.R`
- **Optional merged table:** `make merge` builds `data/processed/merged.csv` for dashboard use; this post joins the same sources directly.
- **Render:** `quarto render stats` from repository root
